## Technical Description

Data for this project comes from the Electric Vehicle Population Data, a
publicly available dataset from the Washington State Department of
Licensing. This is a huge set of data about the EV registered with
Washington State and has the following information:

Vehicle - make and model Type - Battery Electric Vehicle or Plug-in
Hybrid Electric Vehicle Electric Range Base MSRP\
Geography - County City of registration

The data set is a treasure trove of information on the EV diffusion
processes in one of the leading regions of the United States.

Data Collection and Format This information has been collected from
lists of vehicles that the Washington State Department of Licensing
maintains. Data is administratively collected through registration and
updating vehicle lists. The dataset is in CSV format and presents
210,165 records and 17 variables. The presented dataset combines
numeric, categorical, and geographic variables.

Update Frequency While frequency is not directly stated, all vehicle
registration datasets do naturally update with time, as well as with the
passing of ownership and other changes.

## Missing value analysis

Known Issues and Limitations The following are issues that were noted
during exploration:

Missing Values: Some missing values do exist in a few columns, like
Legislative District, Vehicle Location, and Electric Range. This reduces
the scope when analyzing some of these variables.

Geographical scope: The given dataset doesn't enable the view on
nationwide or global views of how electric vehicle adoption is going on;
it's restricted to Washington State only.

Outdated records: Records may not get updated often, and recent
registrations or changes may not be reflected.

Data Cleaning: Categorical values or geographic identifiers may have
inconsistencies that will need cleaning.

Data Import Plan: This is freely downloadable from the Data.gov catalog,
in CSV format. First, this data was imported into this project using
Python, where preliminary cleaning and some exploration had been done.
Later, it would be loaded into R for analysis and visualization. This
dataset provides a robust foundation for analyzing EV adoption and
identifying trends, despite its geographic limitations and minor data
quality issues.

```{r}
install.packages("dplyr") 
install.packages("tidyr")
library(ggplot2)
library(dplyr)
library(tidyr)

ev_data <- read.csv("Electric_Vehicle_Population_Data.csv")

# Calculate the number of missing values for each column
missing_data <- ev_data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "MissingValues")

# Create a bar chart with colored bars
ggplot(missing_data, aes(x = reorder(Variable, -MissingValues), y = MissingValues, fill = MissingValues)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "Missing Values by Variable",
       x = "Variables",
       y = "Number of Missing Values") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Bar Graph: A graph representing the percent of missed values within the
columns; the worst hit is found in the legislative District. The above
bar plot quantitatively views the missing data in terms of their
percentages. The highest rate of missing value is in the Legislative
District column which is about 0.21%, hence the most affected variable
in the dataset. Other columns include Vehicle Location, Electric Range,
and Base MSRP. These all contain very few missing values, under 0.01%,
and are relatively complete, probably without a huge effect on the
analysis. Overall, this data is pretty complete; there are only a few
columns in the entire set with missing information.

```{r}
library(reshape2)

# Create a binary missing value matrix (1 for missing, 0 for not missing)
missing_matrix <- as.data.frame(is.na(ev_data))
missing_matrix <- cbind(Row = 1:nrow(missing_matrix), missing_matrix)

# Melt the data for ggplot
missing_long <- melt(missing_matrix, id.vars = "Row")

# Create the heatmap
ggplot(missing_long, aes(x = variable, y = Row, fill = value)) +
  geom_tile() +
  scale_fill_manual(values = c("red", "black"), labels = c("Not Missing", "Missing")) +
  labs(title = "Heatmap of Missing Values",
       x = "Variables",
       y = "Rows",
       fill = "Missing") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

The following heatmap shows the pattern of missing values across the
dataset. Each vertical band here represents a column, with highlighted
areas showing missing values. Most of the rows are fully filled, and the
rest hold few missing values. There is no indication that the missing
values clustered in rows, which therefore are most likely to be random
and not related to some particular sub-set of the data. The columns most
affected-only showing continuous gaps-include the Legislative District.
This heat map confirms that the missing data points are not pervasive
and affect only a small subset of the data.